---
title: "In_class_Ex05"
author: "Wan Kee"
date: "16 December 2023"
date modified: "13 December 2023"
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: source
---

# 1 Overview

Learning Objectives:

# 2 Load packages

`spflow` is an estimation of spatial econometric models of origin-destination flows.

We will install the development version of `spflow` from github. Go to `Packages` and check the version number is 0.1.0.9010.

```{r}
# devtools::install_github("LukeCe/spflow")
```

`sf` handles spatial data.
`tidyverse` handles attribute data.
`knitr` creates html tables.

```{r}
pacman:::p_load(sf, sp, spdep, tmap, tidyverse, Matrix, knitr)
```

Note: `MASS` is replaced by `Matrix` to speed up computations.

# 3 Prepare data

Three datasets are required to calibrate Spatial Econometric Interaction Models:
1. Spatial weights
2. Tibble data.frame of origins, destination, flows and distance between orign and destination
3. Tibble data.frame of explanatory variables
(Compared to SIM, SEIM does not require OD specification.)

# 3 Import data

:::panel-tabset

# Subzone

Build geographical area

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MPSZ-2019") %>% 
  st_transform(crs=3414)

glimpse(mpsz)
```


```{r}
#mpsz$BUSSTOP_COUNT <- mpsz %>% 
#  st_interacts(mpsz, busstop)
```


```{r}
#busstop_mpsz <- st_intersection(busstop, mpsz) %>%
#  select(BUS_STOP_N, SUBZONE_C) 

#glimpse(busstop_mpsz)
```


```{r}
#mpsz_busstop <- mpsz %>% 
#  filter()
```

:::

```{r}
#centroids <- suppressWarnings({
#  st_point_on_surface(st_geometry(mpsz_busstop))
#})
```


Create a list of spatial weight matrix.

Error in poly2nb(busstop_mpsz) : Polygon geometries required
```{r}
#mpsz_nb <- list(
#  "by_contiguity" = poly2nb(mpsz_busstop),
#  "by_distance" = dnearneigh(centroids,
#                             d1 = 0,
#                             d2 = 5000),
#  "by_knn" = knn2nb(knearneigh(centroids, 3))
#)
```


