---
title: "Geographically weighted regression"
author: "Wan Kee"
date: "30 November 2023"
date modified: "13 December 2023"
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: source
---

# 1. Overview

**Geographically weighted regression** is a spatial statistical technique that takes non-stationary variables into consideration (e.g., climate; demographic factors; physical environment characteristics) and models the local relationships between these independent variables and an outcome of interest (also known as dependent variable). 

Learning Objectives:
1. Build hedonic pricing models
2. Calibrate hedonic pricing models by using GWR methods

# 2. Load packages

The R packages needed for this exercise are as follows:

`olsrr` builds OLS and performing diagnostics tests.
`GWmodel` calibrate geographical weighted family of models.
`corrplot` enables multivariate data visualisation and analysis.
`sf` handles spatial data.
`tidyverse` handles attribute data.
`tmap` plots choropleth maps.

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary)
```

::: {.callout-note}
GWmodel package provides a collection of localised spatial statistical methods, namely: GW summary statistics, GW principal components analysis, GW discriminant analysis and various forms of GW regression; some of which are provided in basic and robust (outlier resistant) forms. Commonly, outputs or parameters of the GWmodel are mapped to provide a useful exploratory tool, which can often precede (and direct) a more traditional or sophisticated statistical analysis.
:::

# 3. Import data

:::panel-tabset
## Subzone

`mpsz` is is a **geospatial** dataset of the Master Plan 2014, a forward looking guiding plan for Singapore's development in the medium term over the next 10 to 15 years published in **2014**.

The output indicates that the geospatial objects are **multipolygon** features. There are **323 features** and 6 fields. It is in **SVY21** projected coordinates system with **XY** dimension.

Source: URA (Download [here](https://beta.data.gov.sg/collections/1708/view))

```{r}
mpsz = st_read(dsn="data/geospatial", layer="MP14_SUBZONE_WEB_PL")
```
## Condo

```{r}
condo_resale <- read_csv("data/aspatial/Condo_resale_2015.csv")
glimpse(condo_resale)
```
:::

# 4. Prepare data

:::panel-tabset

## mpsz

Update `mpsz` with the correct ESPG code 3414

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
```

Verify ESPG code of `mpsz`

```{r}
st_crs(mpsz_svy21)
```


```{r}
st_bbox(mpsz_svy21)
```

## Condo

Convert `condo_resale` data frame into a simple feature data frame by using st_as_sf() of sf package

`st_transform()` of sf package converts the coordinates from WGS84 (crs:4326) to SVY21 (crs=3414).

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

```{r}
head(condo_resale.sf)
```

:::

# 4. Explore data

Plot the distribution of SELLING_PRICE

```{r}
ggplot(data=condo_resale.sf, aes(x=`SELLING_PRICE`)) +
  geom_histogram(bins=20, color="black", fill="light blue")
```


# 5. Plot data

