---
title: "SIMs"
author: "Wan Kee"
date: "9 December 2023"
date modified: "9 December 2023"
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: source
---

The analysis involves the following packages:

-   `sf` imports and handles geospatial data
-   `DT` enables R data objects (matrices or data frames) to be displayed as tables on HTML pages.
-   `tidyverse` performs aspatial data import, wrangling and visualization
-   `spdep` compute spatial weights and spatially lagged variables
-   `spfep` compute spatial autocorrelation
-   `mapview` and `tmap` supports data visualisation

```{r}
pacman::p_load(sf, sp, DT, gridExtra, knitr, mapview, spdep, sfdep, tmap, tidyverse)
```

```{r}
flow_data <- read_rds("data/rds/flow_data_tidy.rds")
glimpse(flow_data)
```

```{r}
flow_data$FlowNoIntra <- ifelse(
  flow_data$ORIGIN_SZ == flow_data$DESTIN_SZ,
  0,
  flow_data$MORNING_PEAK)

flow_data$offset <- ifelse(
  flow_data$ORIGIN_SZ == flow_data$DESTIN_SZ,
  0.000001,
  1)
```


```{r}
inter_zonal_flow <- flow_data %>% 
  filter(FlowNoIntra>0)
```

log(DIST) must be negative where people are not willing to travel to show distance decay.
log(SCHOOL_COUNT) and log(RETAIL_COUNT) should be positive and p-value less than 0.05 to be statistically significant.

Goodness-of-fit

Define a function to calculate R2.

```{r}
CalcRSquared = function(observed, estimated){
  r = cor(observed, estimated)
  R2 = r^2
  R2
}
```


Doubly constrained does not have -1 for log(DIST) where there is no interaction.