---
title: "Spatial Interaction Models"
author: "Wan Kee"
date: "21 November 2023"
date modified: "21 November 2023"
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: source
---

# 1.1 Overview

**Spatial interaction model** is a mathematical model that predicts the movement of people between origins and destinations by examining the distance between them. It is based on (1) their **spatial separation**; (2) their **complementarity**; and (3) other **intervening opportunities** or spatial structural elements serve to augment or diminish the expected flow.

(1) **Spatial separation** is measured by cost or time involved in the interaction, and is often represented by a distance-decay function that shows diminishing interaction as distance increases.

(2) **Complementarity** is an indicator of the suitability of the particular pair of places for the movement 

(3) Interactions may be diverted to **intervening opportunities** with considerations on the relative location of the interacting agents vis-à-vis other places.

Notes diagonals of distance matrix will be zero.

# 1.2 Load packages

```{r}
pacman::p_load(tmap, sf, sp, DT, stplanr, performance, reshape2, ggpubr, units, tidyverse)
```

# 1.3 Import data

mpsz

```{r}
mpsz = st_read(dsn="data/geospatial", layer="MPSZ-2019") %>% 
  st_transform(crs=3414)
```

```{r}
write_rds(mpsz, "data/rds/mpsz.rds")
mpsz <- read_rds("data/rds/mpsz.rds")
mpsz
```

# 1.4 Prepare data

Converting from sf data.table to SpatialPolygonsDataFrame

`as.Spatial()` will be used to convert mpsz from sf tibble data frame to SpatialPolygonsDataFrame of sp object.

```{r}
mpsz_sp <- as(mpsz, "Spatial")
#mpsz_sp <- mpsz %>% 
#  as_Spatial()
mpsz_sp
```

Compute the distance matrix

`spDists()` of sp package will be used to compute the Euclidean distance between the centroids of the planning subzones

```{r}
dist <- spDists(mpsz_sp, longlat = FALSE)
head(dist, n = c(5, 5))
```

there are 332 features hence 332 points, we will limit to 5 columns and 5 rows.

usual: mpsz_sp_selected \<- mpsz_sp %\>% slect(mpsz\@data\$SUBZONE)

sp: no need

Labelling column and row heanders of a distance matrix

Create a list sorted according to the the distance matrix by planning sub-zone code.

```{r}
sz_names <- mpsz$SUBZONE_C
glimpse(sz_names)
```

attach SUBZONE_C to row and column for distance matrix matching ahead

`paste0` concatenates vectors after converting to characters.

```{r}
colnames(dist) <- paste0(sz_names)
rownames(dist) <- paste0(sz_names)
```

Pivoting distance value by SUBZONE_C

`melt()` convert object into molten dataframe.

NEVER SORT THE DATA; SEQUENCE IS IMPT

```{r}
distPair <- melt(dist) %>% 
  rename(dist = value)
head(distPair, 5)
```

Updating intra-zonal distances

append a constant value to replace the intra-zonal distance of 0.

select and find out the minimum value of the distance by using summary().

```{r}
distPair %>% 
  filter(dist > 0) %>% 
  summary()
```

A constant distance value of 50m is added into intra-zones distance.

```{r}
distPair$dist <- ifelse(distPair$dist == 0,
                        50, distPair$dist)
```

rename the origin and destination fields

```{r}
distPair <- distPair %>% 
  rename(orig = Var1,
         dest = Var2)
summary(distPair)
write_rds(distPair, "data/rds/distPair.rds")
```

Prepare data flow

import od_data

```{r}
od_data <- read_rds("data/rds/od_data.rds")
```

compute the total passenger trip between and within planning subzones

```{r}
flow_data <- od_data %>% 
  group_by(ORIGIN_BS, DESTIN_BS) %>% 
  summarise(TRIPS = sum(MORNING_PEAK))
head(flow_data, n = 5)
```

Separating intra-flow from passenger volume df

Add two new fields in flow_data dataframe.

convert data value type of ORIGIN_SZ and DESTIN_SZ fields of flow_data dataframe into factor data type.

```{r}
flow_data$FlowNoIntra <- ifelse(
  flow_data$ORIGIN_BS == flow_data$DESTIN_BS, 
  0, flow_data$TRIPS)
flow_data$offset <- ifelse(
  flow_data$ORIGIN_BS == flow_data$DESTIN_BS, 
  0.000001, 1)

flow_data$ORIGIN_BS <- as.factor(flow_data$ORIGIN_BS)
flow_data$DESTIN_BS <- as.factor(flow_data$DESTIN_BS)

head(flow_data, n = 5)
```

Combining passenger volume data with distance value

left_join() of dplyr will be used to flow_data dataframe and distPair dataframe.

```{r}
flow_data1 <- flow_data %>% 
  left_join(distPair,
            by = c("ORIGIN_BS" = "orig",
                   "DESTIN_BS" = "dest"))
glimpse(flow_data1)
```

Preparing Origin and Destination Attributes

```{r}
pop <- read_csv("data/aspatial/pop.csv")
glimpse(pop)
```

Geospatial data wrangling

Add SZ to pop

```{r}
pop <- pop %>%
  left_join(mpsz,
            by = c("PA" = "PLN_AREA_N",
                   "SZ" = "SUBZONE_N")) %>%
  select(1:6) %>%
  rename(SZ_NAME = SZ,
         SZ = SUBZONE_C)
glimpse(pop)
```

```{r}
SIM_data <- read_rds("data/rds/SIM_data.rds")
glimpse(SIM_data)
```

# 1.4 Explore data

# 1.5 Plot data

Calibrating Spatial Interaction Models

Visualising the dependent variable

plot the distribution of the dependent variable (i.e. TRIPS) by using histogram

```{r}
ggplot(data = SIM_data, aes(x = TRIPS)) +
  geom_histogram(fill = "pink", color = "black",
                 binwidth = 2500) +
  theme(
    panel.background = element_blank()
  )
```

isualise the relation between the dependent variable and one of the key independent variable in Spatial Interaction Model, namely distance.

```{r}
ggplot(data = SIM_data,
       aes(x = dist,
           y = TRIPS)) +
  geom_point(size = 1, colour = "lightblue") +
  geom_smooth(method = lm) +
  theme(
    panel.background = element_blank()
  )
```

plot the scatter plot by using the log transformed version of both variables, we can see that their relationship is more resemble linear relationship.

```{r}
ggplot(data = SIM_data,
       aes(x = log(dist),
           y = log(TRIPS))) +
  geom_point(size = 1, colour = "lightblue") +
  labs(title = "Trips by Distance") +
  geom_smooth(method = lm) + 
  theme(
    panel.background = element_blank()
  )
```
Checking for variables with zero values

```{r}
summary(SIM_data)
```

replace zero values to 0.99

```{r}
SIM_data$DESTIN_AGE7_12 <- ifelse(
  SIM_data$DESTIN_AGE7_12 == 0,
  0.99, SIM_data$DESTIN_AGE7_12)
SIM_data$DESTIN_AGE13_24 <- ifelse(
  SIM_data$DESTIN_AGE13_24 == 0,
  0.99, SIM_data$DESTIN_AGE13_24)
SIM_data$DESTIN_AGE25_64 <- ifelse(
  SIM_data$DESTIN_AGE25_64 == 0,
  0.99, SIM_data$DESTIN_AGE25_64)
SIM_data$ORIGIN_AGE7_12 <- ifelse(
  SIM_data$ORIGIN_AGE7_12 == 0,
  0.99, SIM_data$ORIGIN_AGE7_12)
SIM_data$ORIGIN_AGE13_24 <- ifelse(
  SIM_data$ORIGIN_AGE13_24 == 0,
  0.99, SIM_data$ORIGIN_AGE13_24)
SIM_data$ORIGIN_AGE25_64 <- ifelse(
  SIM_data$ORIGIN_AGE25_64 == 0,
  0.99, SIM_data$ORIGIN_AGE25_64)

summary(SIM_data)
```


Unconstrained Spatial Interaction Model

```{r}
uncSIM <- glm(formula = TRIPS ~ 
                log(ORIGIN_AGE25_64) + 
                log(DESTIN_AGE25_64) +
                log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
uncSIM
```


R-squared function


```{r}
CalcRSquared <- function(observed,estimated){
  r <- cor(observed,estimated)
  R2 <- r^2
  R2
}
```

compute the R-squared of the unconstrained SIM

```{r}
CalcRSquared(uncSIM$data$TRIPS, uncSIM$fitted.values)
```


```{r}
r2_mcfadden(uncSIM)
```

Origin (Production) constrained SIM

```{r}
orcSIM <- glm(formula = TRIPS ~ 
                 ORIGIN_SZ +
                 log(DESTIN_AGE25_64) +
                 log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
summary(orcSIM)
```

examine how the constraints hold for destinations this time.

```{r}
CalcRSquared(orcSIM$data$TRIPS, orcSIM$fitted.values)
```

Destination constrained

```{r}
decSIM <- glm(formula = TRIPS ~ 
                DESTIN_SZ + 
                log(ORIGIN_AGE25_64) + 
                log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
summary(decSIM)
```


```{r}
CalcRSquared(decSIM$data$TRIPS, decSIM$fitted.values)
```


Doubly constrained

```{r}
dbcSIM <- glm(formula = TRIPS ~ 
                ORIGIN_SZ + 
                DESTIN_SZ + 
                log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
summary(dbcSIM)
```

```{r}
CalcRSquared(dbcSIM$data$TRIPS, dbcSIM$fitted.values)
```

